---
title: "Random generation: Inverse method. Examples "
author: 'Arnau Pérez Reverte'
output:
  html_document:
    code_folding: show
    highlight: haddock
    keep_md: yes
    number_sections: yes
    theme: readable
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  pdf_document:
    toc: yes
---
# Statement: Random generation Inverse method: Exponential

```{r}

#tini <- proc.time()

lambda <- 2
nsim <- 10^5
set.seed(1)
U <- runif(nsim)
random_values <- -log(U)/lambda # -log(1-U)/lambda
#tiempo <- proc.time() - tini
#tiempo
## user system elapsed
## 0.00 0.01 0.02
hist(random_values, breaks = "FD", freq = FALSE, main= "Exponential Random generation",
xlim = c(0, 5), ylim = c(0, 2.5), col="blue")
curve(dexp(x, lambda), lwd = 2, add = TRUE, col="green")

```

# Statement: Random generation Inverse method: Weibull
What models
## Aims and Scopes
Parameters, hazar function, risk function
## Relationship with other models
## Generator
```{r}

#tini <- proc.time()

lambda <- 2
k <- 10
nsim <- 10^5
set.seed(1)
U <- runif(nsim)
random_values <- lambda*(-log(U))^(1/k)
#tiempo <- proc.time() - tini
#tiempo
## user system elapsed
## 0.00 0.01 0.02
hist(random_values, breaks = "FD", freq = FALSE, main= "Weibull Random generation",
xlim = c(0, 5), ylim = c(0, 2.5), col="blue")
curve(dweibull(x, k, lambda), lwd = 2, add = TRUE, col="green")

```

# Statement: Random generation Inverse method: Cauchy

## Aims and Scopes
## Relationship with other models
## Generator
```{r}
#tini <- proc.time()

gam <- 0.5
x0 <- 0
nsim <- 10^5
set.seed(1)
U <- runif(nsim)
random_values <- x0+gam*tan(pi*(U-1/2))

hist(random_values, 
     breaks = "FD", 
     freq = FALSE, 
     main= "Cauchy Random generation", 
     xlim = c(-5, 5), 
     ylim = c(0, 2.5), 
     col="blue")
curve(dcauchy(x, location = x0, scale = gam), lwd = 2, add = TRUE, col="green")
```
# Statement: Random generation Inverse method: Gamma

## Aims and Scopes
## Relationship with other models
## Generator


# Statement: Random generation Inverse method: Gamma

## Aims and Scopes
## Relationship with other models
## Generator
